 

### 2021-06-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Interestingly, within-canopy Japanese oak sun and shade leaves were found to maintain similar rates of excess energy production under respective peak light conditions. Excessive light energy not utilized can cause photoinhibition. Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain higher photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018]. Perhaps, differences in these thermal stablization strategies realize similar $T_{opt}$ for photosynthesis in sun and shade leaves.

@@

In 1994, Knapp et al. showed that $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions. In both species $T_{50}$ also increased from June to August. Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.

@@

Biophysical conditions can alter growth rates, but can do so differently between size classes through influencing existing demographic differences. Particulary evidence in in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014]. Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al. 2007*], perhaps due to increase in nighttime metabolic respiration that could supress growth.

**Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].

>>>>

@hernandezSimilarTemperatureDependence2020 speculated that the lack of sun-shade differences in Topt may reflect acclimation to similar temperatures of peak photosynthesis. Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of mid-day stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous. Shade leaves that do not experience mid-day depression continue to fix carbon even when afternoon air temperatures peak. Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.

@@

In 1994, Knapp et al. showed that $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions. In both species $T_{50}$ also tracked increased from June to August. Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.

@@

Biophysical conditions can alter growth rates, but can do so differently between size classes through influencing existing demographic differences. Particulary evidence in in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014]. Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al. 2007*], perhaps due to increase in nighttime respiratory carbon loss that reduced carbon availability for growth.

**Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with increasing $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
